---
title: "Problem set 2"
author: "David Segovia"
output: html_document
---
   
   
   

```{r}
library(tidyverse)
setwd("~/Downloads/PA 528 Public Program Evaluation/Problem Sets/PS2/data")
car <- read_csv("car_data.csv")


```


# Learning R
   
Tell me that you worked through the primers and videos and examples at the example page for this week:

I did all the primers and had the time of my life!
   




# My first plots

Insert a chunk below and use it to create a scatterplot (hint: `geom_point()`) with diplacement (`displ`) on the x-axis, city MPG (`cty`) on the y-axis, and with the points colored by drive (`drv`).

```{r}

ggplot(data= car) + 
   geom_point(mapping = aes(x= displ, y= cty, color= drv))



```




Insert a chunk below and use it to create a histogram (hint: `geom_histogram()`) with highway MPG (`hwy`) on the x-axis. Do not include anything on the y-axis (`geom_histogram()` will do that automatically for you). Choose an appropriate bin width. If you're brave, facet by drive (`drv`).


```{r}

ggplot(data = car) + 
   geom_histogram(mapping =aes(x= hwy), binwidth = 1) + facet_wrap(~drv)



```
# My first data manipulation

Insert a chunk below and use it to calculate the average city MPG (`cty`) by class of car (`class`). This won't be a plot---it'll be a table. Hint: use a combination of `group_by()` and `summarize()`.


```{r}
car %>%
   group_by(class) %>%
   summarise(averagecity= mean(cty))


```




























