---
title: "Problem set 5: The diff-in-diff"
author: "David Segovia"
date: "Nov 5 2021"
output: 
  html_document: 
    toc: yes
  pdf_document: 
    latex_engine: xelatex
    toc: yes
  word_document: 
    toc: yes
---

---

In 1996, Nada Eissa and Jeffrey B. Liebman published a study on the effect of the Earned Income Tax Credit (EITC) on employment (http://darp.lse.ac.uk/papersdb/Eissa-Liebman_(QJE96).pdf). The EITC is a special tax credit for low income workers that changes depending on (1) how much a family earns (the lowest earners and highest earners don't receive a huge credit, as the amount received phases in and out), and (2) the number of children a family has (more kids = higher credit). See a brief explanation of EITC policy and how the EITC works from Center on Budget and Policy Priorities (https://www.cbpp.org/research/federal-tax/policy-basics-the-earned-income-tax-credit).

 
Eissa and Liebman's study looked at the effects of the EITC on women's employment and wages after it was initially substantially expanded in 1986. The credit was expanded substantially again in 1993. For this problem set, you'll measure the causal effect of this 1993 expansion on the employment levels and annual income for women.


A family must have children in order to quality for the EITC, which means the presence of 1 or more kids in a family assigns low-income families to the EITC program (or "treatment"). Those families without kids are not eligible for the EITC program ("control"). We have annual data on earnings from 1991–1996, and because the expansion of EITC occurred in 1993, we also have data both before and after the expansion. This treatment/control before/after situation allows us to use a difference-in-differences approach to identify the causal effect of the EITC.

The dataset provided (`eitc.dta`) is a Stata data file containing more than 13,000 observations. This is non-experimental data—the data comes from the US Census's Current Population Survey (CPS) and includes all women in the CPS sample between the ages of 20–54 with less than a high school education between 1991–1996. There are 11 variables:

- `work`: Dummy variable indicating if the woman was employed last year (1 = employed)
- `state`: The woman's state of residence. The numbers are Census/CPS state numbers: [http://unionstats.gsu.edu/State_Code.htm](http://unionstats.gsu.edu/State_Code.htm)
- `year`: The tax year
- `urate`: The unemployment rate in the woman's state of residence
- `children`: The number of children the woman has
- `nonwhite`: Binary variable indicating if the woman is not white (1 = Hispanic/Black)
- `finc`: The woman's family income in 1997 dollars
- `earn`: The woman's personal income in 1997 dollars
- `age`: The woman's age
- `ed`: The number of years of education the woman has
- `unearn`: The woman's family income minus her personal income, in *thousands* of 1997 dollars






```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)  # For ggplot, %>%, mutate, filter, group_by, and friends
library(haven)      # For loading data from Stata
library(broom)      # For showing models as data frames
setwd("~/Library/Mobile Documents/com~apple~CloudDocs/PA 528 Public Program Evaluation/Class 11")
# Load EITC data
eitc <- read_stata("eitc.dta") %>% 
  # case_when() is a fancy version of ifelse() that takes multiple conditions
  # and outcomes. Here, we make a new variable named children_cat(categorical) 
  # with three different levels: 0, 1, and 2+
  mutate(children_cat = case_when(
    children == 0 ~ "0",
    children == 1 ~ "1",
    children >= 2 ~ "2+"
  ))


```

# 1. Exploratory data analysis

Create a new variable that shows if women have 0 children, 1 child, or 2+ children (This has been provided above). 

What is the average of `work`, `finc`, `earn`, `nonwhite`, `ed`, and `age` across each of these different levels of children? How are these groups different? Describe your findings in a paragraph.

## Work



```{r}
# Work
eitc %>% 
  group_by(children_cat) %>% 
  summarize(avg_work = mean(work))

# stat_summary() here is a little different from the geom_*() layers you've seen
# in the past. stat_summary() takes a function (here mean_se()) and runs it on
# each of the children_cat groups to get the average and standard error. It then
# plots those with geom_pointrange. The fun.args part of this lets us pass an
# argument to mean_se() so that we can multiply the standard error by 1.96,
# giving us the 95% confidence interval
ggplot(eitc, aes(x = children_cat, y = work)) +
  stat_summary(geom = "pointrange", fun.data = "mean_se", fun.args = list(mult = 1.96))

```

## Family income

```{r}
eitc %>% 
  group_by(children_cat) %>% 
  summarize(avg_inc = mean(finc))

ggplot(eitc, aes(x = children_cat, y = finc)) +
  stat_summary(geom = "pointrange", fun.data = "mean_se", fun.args = list(mult = 1.96))

```


## Earnings 

```{r}

eitc %>% 
  group_by(children_cat) %>% 
  summarize(avg_earn = mean(earn))

ggplot(eitc, aes(x = children_cat, y = earn)) +
  stat_summary(geom = "pointrange", fun.data = "mean_se", fun.args = list(mult = 1.96))

```


## Race



```{r}

eitc %>% 
  group_by(children_cat) %>% 
  summarize(avg_nonwhite = mean(nonwhite))
# remember, non-white is binary variable indicating if the person is non-white(Hispanic or Black) or not. So 1= Hispanic/Black, 0 means white.
# It seems the highest proportion of Black/Hispanic women have 2 or more kids compared to white women. These levels drop down when looking at only 1 kid, where Hispanic/Black women still have a little bit more than average. But the levels are almost the same for women with 0 kids between white and non-white. 

ggplot(eitc, aes(x = children_cat, y = nonwhite)) +
  stat_summary(geom = "pointrange", fun.data = "mean_se", fun.args = list(mult = 1.96))

```


## Education

```{r}

eitc %>% 
  group_by(children_cat) %>% 
  summarize(avg_education = mean(ed))

ggplot(eitc, aes(x = children_cat, y = ed)) +
  stat_summary(geom = "pointrange", fun.data = "mean_se", fun.args = list(mult = 1.96))

```


## Age

```{r}
eitc %>% 
  group_by(children_cat) %>% 
  summarize(avg_age = mean(age))

ggplot(eitc, aes(x = children_cat, y = age)) +
  stat_summary(geom = "pointrange", fun.data = "mean_se", fun.args = list(mult = 1.96))


```


## General summary

**Describe your findings in a paragraph. How do these women differ depending on the number of kids they have?**

Women without kids are more likely to work (0.57) than women who have 1 (0.53) or 2+ kids (0.42). Women without kids are also more likely to have higher average incomes (18,559) compared to women with 1 (13,941) and 2+ kids (11,985) and higher earnings (13,760 compared to 9,928 and 6,613 respectively). 

Women who are Black or Hispanic are also more than likely than white women to have 2 or more kids(0.70) and a little more likely to have 1 kid (0.59). However, race does not seem to play a critical factor to whether women have no kids since the proportion is almost 50-50 (0.51). 

Women with more kids are also more likely to be educated than women without kids (9 years compared to 8.5 years). 

Surprisingly, women with 2 or more kids are, on average, around 32 years old. These women are a little younger compared to women with only 1 kid, who are around ~33 years old. However, woman without kids are a little older, around ~38 years old.



# 2. Create treatment variables

Create a new variable for treatment named `any_kids` (should be TRUE or 1 if `children` > 0) and a variable for the timing named `after_1993` (should be TRUE or 1 if `year` > 1993).

Remember you can use the following syntax for creating a new binary variable based on a test:

```r
new_dataset <- original_dataset %>%
  mutate(new_variable = some_column > some_number)
```

```{r}

# Make new dataset here. You can either do something like:

eitc_a <- eitc %>% 
  mutate(any_kids = case_when(
    children == 0 ~ "0",
    children > 0 ~ "1"
    ))

eitc_b <-eitc_a %>%
  mutate(after_1993 = case_when(
    year > 1993 ~ "1",
    year <= 1993 ~ "0"
  ))


```


# 3. Check pre- and post-treatment trends

Create a new dataset that shows the average proportion of employed women (`work`) for every year in both the treatment and control groups (i.e. both with and without kids). (Hint: use `group_by()` and `summarize()`, and group by both `year` and `any_kids`.)

```{r}
# Find average of work across year and any_kids

a <- eitc_b %>% 
  group_by(year, any_kids) %>%
  summarize(mean_work = mean(work))
a

```

Plot these trends using colored lines and points, with year on the x-axis, average employment on the y-axis. Add a vertical line at 1994 (hint: use `geom_vline(xintercept = SOMETHING)`. 

```{r}
# Add plot here, with x = year, y = average employment, and color = any_kids.
# Add a vertical line too.

ggplot(a, aes(x = as.factor(year),
                  y = mean_work,
                  color = as.factor(any_kids))) +
  geom_point() +
  geom_line(aes(group = as.factor(any_kids))) +
  geom_vline(aes(xintercept="1994"))

```

**Do the pre-treatment trends appear to be similar?**

Yes, they appear to be similar. Between 1991 and 1992, both groups with kids and without kids had a slight drop in employment, consistent levels from 1992 and 1993, and a slight increase from 1993 to 1994. Overall, it seems like women without any kids are working more than women without kids.



# 4. Difference-in-differences with regression

Run a regression model to find the diff-in-diff estimate of the effect of the EITC on employment (`work`) (hint: remember that you'll be using an interaction term).

```{r}
# Regression model here

model_diff <- lm(work ~ any_kids + after_1993 + any_kids * after_1993,
                  data = eitc_b)
summary(model_diff)
tidy(model_diff)

```


**What is the difference-in-differences estimate? Discuss the result.** (Hint, these numbers are percents, so you can multiply them by 100 to make it easier to interpret. For instance, if the diff-in-diff number is 0.15 (it's not), you could say that the EITC caused the the proportion of mothers in the workplace to increase 15 percentage points.)

**Based on the results, what is the relationship between the problem of EITC and the employment of mothers in the workplace? Does this mean EITC program/policy *cause* employment? Explain.**

The coefficient of the difference-in-differences estimate is 0.047, both positive and significant. This means that the EITC caused the proportion of mothers in the workplace to increase 4.7 percentage points. This is statistically significant (p<.001 **). It seems like the EITC did cause the employment, but we are not controlling for other variables like demographics and education.


# 5. Difference-in-difference with regression and controls

Run a new regression model with demographic controls. Eissa and Liebman used the following in their original study: non-labor income (family income minus personal earnings, or the `unearn` column), number of children, race, age, age squared, education, and education squared. You'll need to make new variables for age squared and education squared. (These are squared because higher values of age and education might have a greater effect: someone with 4 years of education would have 16 squared years, while someone with 8 years (twice as much) would have 64 squared years (way more than twice as much).)

```{r}
# Create new variables for age squared and education squared

eitc_b$age_squared = eitc$age^2
eitc_b$ed_squared = eitc$ed^2
eitc_b %>% select(age, age_squared, ed, ed_squared)

# Regression model with demographic controls here


model_diff2 <- lm(work ~ unearn + children + nonwhite + age + age_squared + ed + ed_squared + any_kids + after_1993 + any_kids * after_1993, data = eitc_b)
summary(model_diff2)

# R tends to put interaction terms last in regression tables, so you might not
# see the any_kids * after_1993 coefficient on the first page of the table here




```

**Does the treatment effect change? Interpret these findings.**


After controlling for other variables, difference-in-difference coefficient did change a little bit. The EITC program effect went from 4.7 to 5.8, a difference of 1.1 after controlling for additional variables.  The interpretation in this model is that after controlling for demographic variables, the EITC program increased the proportion of women in the workplace by ~5.8 percentage points. This is also statistically significant at the .001*** level.
